Parallel Programming in Haskell Almost for Free

### 2013-09-30 src2/embarbb/implementation.tex
apa

\section{Implementation} 
EmbArBB is a deeply embedded language. There is a constructor for 
every operation that ArBB can perform. A deep embedding is useful when 
there is need to apply optimisations or transformations to the AST before 
executing the operations. However, EmbArBB currently relies entirely on 
ArBB to perform optimisations to the program. The only optimisation 
performed on the EmbArBB side is sharing detection. Detecting sharing 
reduces the number of calls into the ArBB C library. Should we add a 
GPU backend to EmbArBB, further GPU specific optimisations of the AST 
will likely become necessary.

\subsection{Vectors} 
EmbArBB has support for one, two and three dimensional dense vectors,
represented by a datatype called {\tt DVector}.

@@

\subsection{The language} 
The EmbArBB language is implemented as a set of library functions, operating on an expression datatype:

@@

\subsection{Interfacing with Haskell and Code generation}

>>>>

\subsection{Implementation} 
EmbArBB is a deeply embedded language. There is a constructor for 
every operation that ArBB can perform. A deep embedding is useful when 
there is need to apply optimisations or transformations to the AST before 
executing the operations. However, EmbArBB currently relies entirely on 
ArBB to perform optimisations to the program. The only optimisation 
performed on the EmbArBB side is sharing detection. Detecting sharing 
reduces the number of calls into the ArBB C library. Should we add a 
GPU backend to EmbArBB, further GPU specific optimisations of the AST 
will likely become necessary.

\subsubsection{Vectors} 
EmbArBB has support for one, two and three dimensional dense vectors,
represented by a datatype called {\tt DVector}.

@@

\subsubsection{The language} 
The EmbArBB language is implemented as a set of library functions, operating on an expression datatype:

@@

\subsubsection{Interfacing with Haskell and Code generation}

### 2013-09-30 src2/embarbb/programming.tex
apa

\section{Programming in ArBB}
\label{sec:ProgrammingArBB}

@@

%\FloatBarrier
\section{Programming in EmbArBB} 
\label{sec:Programming}

@@

% \FloatBarrier 
\subsection{Saxpy} 
{\em Saxpy} is a vector operation that gets its name from the 
description of what it does ``{\em S}ingle-precision {\em A}lpha {\em X} {\em P}lus {\em Y}''. 
There is also a double-precision version called {\em daxpy}. Here we can use a single source 
to obtain both versions:
%. The code listing below {\em axpy} illustrates this:

@@

\subsection{Matrix-vector multiplication} 
In section~\ref{sec:ProgrammingArBB}, matrix-vector multiplication was shown using the C++ ArBB interface. 
Here, EmbArBB is used to implement the same function. The function definition itself 
is very similar to the C++ one:

@@

\subsection{Matrix-matrix multiplication}

@@

\subsection{Histogram}
\FloatBarrier

@@

\FloatBarrier
\subsection{Sobel edge detection filter}
\label{sec:Sobel}
\FloatBarrier

@@

\subsection{Image filtering} 
\label{sec:Blur}
The blur operation, which is a kind of spatial linear filter, shows another way to implement a stencil operation. The {\tt mapStencil} function used is a composite 
function in the EmbArBB library; it is not provided as a primitive from 
the ArBB VM.

@@

\subsection{Sparse matrix multiplication}\label{sec:smvm}
A sparse matrix can be represented by three vectors, one (cidx) containing
column indices, one (offsets) indicating indices in the vector of values
at which the first non-zero element of each row appears, and one containing the
values themselves. This is known as the Compressed Sparse Row (CSR) format.
Given such a matrix and a dense vector, the relevant elements of the dense vector
can be gathered into a new vector using the column indices. The resuting vector
is multiplied pointwise by the vector of non-zero values from the sparse matrix. All that remains, then, is to divide the result into rows and sum each one. The offsets vector
points to where the division should happen.
The code in EmbArBB is as follows:

>>>>

\subsection{Programming in ArBB}
\label{sec:ProgrammingArBB}

@@

%\FloatBarrier
\subsection{Programming in EmbArBB} 
\label{sec:Programming}

@@

% \FloatBarrier 
\subsubsection{Saxpy} 
{\em Saxpy} is a vector operation that gets its name from the 
description of what it does ``{\em S}ingle-precision {\em A}lpha {\em X} {\em P}lus {\em Y}''. 
There is also a double-precision version called {\em daxpy}. Here we can use a single source 
to obtain both versions:
%. The code listing below {\em axpy} illustrates this:

@@

\subsubsection{Matrix-vector multiplication} 
In section~\ref{sec:ProgrammingArBB}, matrix-vector multiplication was shown using the C++ ArBB interface. 
Here, EmbArBB is used to implement the same function. The function definition itself 
is very similar to the C++ one:

@@

\subsubsection{Matrix-matrix multiplication}

@@

\subsubsection{Histogram}
\FloatBarrier

@@

\FloatBarrier
\subsubsection{Sobel edge detection filter}
\label{sec:Sobel}
\FloatBarrier

@@

\subsubsection{Image filtering} 
\label{sec:Blur}
The blur operation, which is a kind of spatial linear filter, shows another way to implement a stencil operation. The {\tt mapStencil} function used is a composite 
function in the EmbArBB library; it is not provided as a primitive from 
the ArBB VM.

@@

\subsubsection{Sparse matrix multiplication}\label{sec:smvm}
A sparse matrix can be represented by three vectors, one (cidx) containing
column indices, one (offsets) indicating indices in the vector of values
at which the first non-zero element of each row appears, and one containing the
values themselves. This is known as the Compressed Sparse Row (CSR) format.
Given such a matrix and a dense vector, the relevant elements of the dense vector
can be gathered into a new vector using the column indices. The resuting vector
is multiplied pointwise by the vector of non-zero values from the sparse matrix. All that remains, then, is to divide the result into rows and sum each one. The offsets vector
points to where the division should happen.
The code in EmbArBB is as follows:

### 2013-09-30 src2/embarbb/related.tex
apa

\section{Related Work} 
\subsection{Data.Array.Accelerate} 
Accelerate~\cite{accelerate} is an embedded language for general purpose, 
flat data parallel computations on GPUs.

@@

\subsection{Data Parallel Haskell} 
Data Parallel Haskell (DPH) is an extension to GHC for nested data parallelism. 
This is one thing that DPH and ArBB have in common.

@@

\subsection{Feldspar} 
Feldspar is a language for Digital Signal Processing (DSP) programming
developed at Chalmers, ELTE University and Ericsson 
(the telecom company)~\cite{FELDSPAR2010}. The functional {\tt while} loop 
of EmbArBB is inspired by the same concept in Feldspar.

@@

\subsection{Nikola} 
Nikola is an embedded language for GPU programming\cite{NIKOLA}, also
in Haskell.  
During the early phases of implementation of EmbArBB, the source code of 
Nikola was often studied for inspiration.

@@

\subsection{Repa} 
Repa is a library for regular, shape polymorphic parallel arrays~\cite{REPA}. Repa 
uses a concept of {\tt delayed} arrays to obtain fusion of operations, such as the 
typical map fusion:
\begin{verbatim}
map f . map g  = map (f . g)
\end{verbatim} 
A delayed array 
has a representation that is quite direct to parallelise; it is represented 
as a function from an index-space to an element and the extents of that same 
index-space. Concretely:

@@

\section{Motivation}

>>>>

\subsection{Related Work} 
\subsubsection{Data.Array.Accelerate} 
Accelerate~\cite{accelerate} is an embedded language for general purpose, 
flat data parallel computations on GPUs.

@@

\subsubsection{Data Parallel Haskell} 
Data Parallel Haskell (DPH) is an extension to GHC for nested data parallelism. 
This is one thing that DPH and ArBB have in common.

@@

\subsubsection{Feldspar} 
Feldspar is a language for Digital Signal Processing (DSP) programming
developed at Chalmers, ELTE University and Ericsson 
(the telecom company)~\cite{FELDSPAR2010}. The functional {\tt while} loop 
of EmbArBB is inspired by the same concept in Feldspar.

@@

\subsubsection{Nikola} 
Nikola is an embedded language for GPU programming\cite{NIKOLA}, also
in Haskell.  
During the early phases of implementation of EmbArBB, the source code of 
Nikola was often studied for inspiration.

@@

\subsubsection{Repa} 
Repa is a library for regular, shape polymorphic parallel arrays~\cite{REPA}. Repa 
uses a concept of {\tt delayed} arrays to obtain fusion of operations, such as the 
typical map fusion:
\begin{verbatim}
map f . map g  = map (f . g)
\end{verbatim} 
A delayed array 
has a representation that is quite direct to parallelise; it is represented 
as a function from an index-space to an element and the extents of that same 
index-space. Concretely:

@@

\subsection{Motivation}

### 2013-09-30 src2/embarbb/results.tex
apa

\section{Benchmarks}
\label{sec:Benchmarks}
%\FloatBarrier

@@

\subsection{matrix-matrix multiplication}
The matrix multiplication benchmarks consist of square matrices 
of sizes 256x256, 384x384, 512x512, 640x640 and 768x768. 
Figure~\ref{fig:mmchart1} shows the runtimes for the four implementations 
being compared, using the best settings in numbers of cores or threads found 
by previous experiments. The sequential C++ and ArBB versions come 
directly from the ArBB distribution.

@@

\subsection{Sobel edge detection}
In the Repa Sobel code, only the applications of the stencils are timed. This 
part is defined as follows 
\begin{verbatim}
gradientX :: Monad m => Image -> m Image
gradientX img
        = computeP
        $ forStencil2 (BoundConst 0) img
              [stencil2| -1  0  1
                         -2  0  2
                         -1  0  1 |]

@@

\subsection{Blur}

@@

\subsection{About the numbers} 
This section presents three benchmarks, two of which compare to Repa only and 
one that compares to Repa, ArBB in C++ and sequential C++ code. In all of these comparisons,
JIT compilation time is excluded.

>>>>

\subsection{Benchmarks}
\label{sec:Benchmarks}
%\FloatBarrier

@@

\subsubsection{matrix-matrix multiplication}
The matrix multiplication benchmarks consist of square matrices 
of sizes 256x256, 384x384, 512x512, 640x640 and 768x768. 
Figure~\ref{fig:mmchart1} shows the runtimes for the four implementations 
being compared, using the best settings in numbers of cores or threads found 
by previous experiments. The sequential C++ and ArBB versions come 
directly from the ArBB distribution.

@@

\subsubsection{Sobel edge detection}
In the Repa Sobel code, only the applications of the stencils are timed. This 
part is defined as follows 
\begin{verbatim}
gradientX :: Monad m => Image -> m Image
gradientX img
        = computeP
        $ forStencil2 (BoundConst 0) img
              [stencil2| -1  0  1
                         -2  0  2
                         -1  0  1 |]

@@

\subsubsection{Blur}

@@

\subsubsection{About the numbers} 
This section presents three benchmarks, two of which compare to Repa only and 
one that compares to Repa, ArBB in C++ and sequential C++ code. In all of these comparisons,
JIT compilation time is excluded.
